Rapid response of mesophyll conductance to light availability allows shade leaves to take advantage of sunflecks

### 2015-10-21 manuscript/manuscript_wtc3.Rmd
manuscript edits from MT

\
During A, CO~2~ diffuses from the atmosphere through stomata (g~s~), intercellular air space and into the chloroplast for fixation by Rubisco (mesophyll conductance, g~m~) [@tazoe2011using]. Based on optimality theory, stomatal conductance should be efficiently distributed within a canopy to utilize supplies of light, N and water to maximize A [@peltoniemi2012co]. This is because stomata are hypothesized to exhibit an optimal behaviour to maximize C gain while simultaneously minimizing water loss through transpiration [@cowan1977stomatal]. Mesophyll conductance can also impose limitations on A as large g~s~ [@warren2008stand; @ubierna2011estimation], reducing the efficiency of leaf N use in A [@niinemets2007photosynthesis]. Variation in photosynthetic capacity between sun and shade leaves has been proposed to be due to differences in g~m~ [@piel2002effect; @warren2007internal], yet the trade-offs that constrain this diffusion pathway are yet to be explicitly quantified. Stomatal and mesophyll conductance should not be considered independent of each other [@griffiths2013mesophyll], but a lack empirical data currently hinders are ability to interpreting their coupled responses to A across sun and shade leaves.

\
Additionally, assessing shade leaf behaviour is made difficult with accounting of short term light fluctuations within a canopy, via sunflecks. How shade leaves utilize sunflecks for short term C gain depends on the combined response time of g~s~ and g~m~ and the underlying photosynthetic biochemistry acclimated to a low light environment [@pearcy1990sunflecks;@tausz2005dynamic]. For example, at short timescales the utilization of sunflecks is limited by delayed responses of stomata, effectively limiting the maximum assimilation rate that can be achieved [@vico2011effects; @way2012sunflecks]. Mesophyll conductance has been shown to respond to environmental factors (e.g. CO~2~, temperature or VPD) across short time scales, possibly faster than g~s~ [@flexas2008mesophyll and references therein], yet short term response to light availability are unclear. For example, g~m~ was found to be independent of light intensity in wheat leaves [@tazoe2009light] but was responsive to light in tobacco [@flexas2007rapid]. Anatomical parameters which regulate g~m~ with changing irradiance such as chloroplast surface area [@terashima2006irradiance] and mesophyll thickness [@boardman1977comparative] are also unlikely to adapt during short light fluctuations. The physiological behaviour of shade leaves to maximize C gain must be assessed as both a degree of acclimation to local irradiance and as a potential response to transitory light availability.

@@

\
In this study we use *Eucalyptus tereticornis* Sm. trees, planted in climate controlled whole tree chambers with ambient and elevated temperature (+3&deg;C) treatments, to empirically evaluate the distribution of resources and leaf physiological behaviour of sun and shade leaves. Our hypotheses are as follows:

@@

\
Before seedlings were planted into each chamber they were maintained under well-watered conditions in 35 L pots and kept inside each chamber. This allowed for seedlings to gain sufficient biomass before planting while also allowing them to acclimate to chamber temperature treatments. Seedlings were planted into each chamber after mean seedling height reached 100 cm. The top soils at this site, used in both pots and chambers, are an alluvial formation of low-fertility sandy loam soils with low organic matter and low water holding capacity. A root exclusion barrier extended from chamber walls to the hard layer (ca. 1 m) and roots were allowed to grow freely below the barrier. After 2 months, floors were installed 45 cm from the soil surface, enabling chamber fluxes of CO~2~ and H~2~O from the whole tree canopy to be monitored. Leaf gas exchange measurements were initiated in October 2013 when trees had both ample height growth and canopy development for realistic canopy light gradients to be measured. At this point, trees under AT treatment had a mean diameter of 28.2&pm;1.1 mm, height of 348&pm;15.1 cm and an estimated leaf area of 3.9&pm;0.1 m^2^. For ET treatments, trees had a mean diameter of 34.1&pm;2.1 mm, height of 418.3&pm;23.1 cm and an estimated leaf area of 6.2&pm;0.2 m^2^.

##Leaf gas exchange, online carbon isotope discrimination and mesophyll conductance
Leaf gas exchange measurements were performed in October 2013 and monthly from December 2013 to April 2014. Measurements were taken on a representative sun and shade leaf for each tree during each measurement campaign.  The newest fully expanded leaf from the stem apex was chosen for gas exchange measurements and sun leaves were measured in the upper third of the canopy. Here, shade leaves are defined as inner-canopy leaves developing in a low light environment. In order to minimize any confounding effects of leaf age, shade leaves were measured first in the lower canopy then gradually higher during each campaign. As shade leaves most likely developed slower this assured that older leaves in the lower canopy were avoided.

\
Prior to gas exchange measurements photosynthetic photon flux density (PPFD) was recorded at the individual leaf level and at the canopy position for each selected leaf.  A hand-held quantum sensor (LI-COR, Lincoln, NE, USA) was used to record leaf level PPFD to ensure that chosen leaves were positioned in the desired light environment, either full sun or full shade. A ceptometer (AccuPAR LP-80, Decagon Devices, Pullman, USA) was then used to measure integrated PPFD at the canopy height of each chosen leaf type. Each ceptometer reading integrated an array of 80 sensors over a total length of 84 cm. Five ceptometer readings were recorded within the canopy at the same height and close to each selected leaf. The mean of these readings was assumed to represent the local light environment of sun and shade leaves for each tree. All measurements of PPFD and gas exchange were performed on sunny days between 10:00-14:30 h.

\
Leaf level gas exchange was measured with a standard (2 x 3 cm) leaf chamber using a portable gas exchange system (LI-6400, LI-COR, Lincoln, NE, USA).  This system was coupled with a tunable diode laser (TDL; TGA100,Campbell Scientific, Inc., Logan, UT, USA) for concurrent measurements of online C isotope discrimination. The CO~2~ in the leaf chamber was set at ambient atmospheric [CO~2~] (400 ppm) with a flow rate of 200 $\mu$mols s^-1^. Two identical gas exchanges systems were run simultaneously, one in each of a randomly chosen WTC for each temperature treatment. Leaf temperatures were controlled at the current AT or ET chamber air temperature. PPFD in the chamber was set to match the individual light environment of each leaf type (explained above). Sun flecks were simulated for shade leaves by increasing the chamber PPFD to the light environment of the full sun leaf in the same tree. The maximum sunfleck response was recorded once CO~2~ and water vapor fluxes re-stabilized in the leaf cuvette (ca. 25 min).

\
Once CO~2~ and water vapor flux values were stable for each leaf measurement, the sample and reference gas lines were diverted to the TDL via T-junctions inserted into the reference gas tube and match valve outlet. These gases were dried by passing through napion gas dryers in the respective gas lines, and then ^12^CO~2~ and ^13^CO~2~ concentrations were measured for each gas by the TDL. Reference, sample and 2 calibration gases were run on alternating 80 s loops (20 s each), one for each paired WTC leaf, for a total of 12 min. This allowed for 4-5 measurements per leaf and data were averaged over the last 10 s of reference and samples gases for calculations. The two calibration gases were drawn from compressed air tanks (330 and 740 ppm CO~2~) in order to correct for gain drift of the TDLAS on each measurement cycle. Photosynthesis, g~s~, transpiration, VPD and intercellular [CO~2~] (C~i~) were auto-logged every 15 s for each gas exchange system over the same 12 min interval.

\
Using online C isotope discrimination measurements, the difference  between the observed discrimination and what is predicted for light saturated gas exchange is proportional to g~m~ [@griffiths2013mesophyll]. First, leaf discrimination was calculated by comparing the isotopic composition of the reference gas entering the chamber ($\delta$^13^C~e~) with the sample gas ($\delta$^13^C~o~) such that:

@@

and $\xi$ is the ratio of the CO~2~ entering the well mixed leaf cuvette to the CO~2~ draw down by the leaf.

@@

\
(7)
$$
\begin{aligned}
\Delta_e = &\frac{1+t}{1-t}\left(\frac{eR_D}{(A+R_d)C_a}(C_i-\Gamma_*)\right)
\end{aligned}
$$

@@

where C~s~ is the CO~2~ partial pressure at the leaf surface, a~b~ is the fractionation from boundary layer diffusion (2.9&permil;) and a is the fractionation due to diffusion in air (4.4&permil;) [@evans1986carbon]. C~a~ and C~i~ are the atmospheric and intercellular partial pressures and $\Gamma_{*}$ is the compensation point in the absence of mitochondrial respiration in the light (R~d~). In this experiment both $\Gamma_{*}$ and R~d~ were derived using a standard Arrhenius function with parameters for *Eucalyptus globulis* from [@crous2012light]. The ternary effect corrections (t) are described by:

@@

where C~c~ is the chloroplast CO~2~ partial pressure. Once g~m~ was calculated C~c~ and the drawdown of CO~2~ from the intercellular air spaces to the site of carboxylation were then estimated using Equation 11. Examples of this approach to measure gas exchange and C isotope discrimination are presented in [@evans2013temperature]. The variation in $\Delta$~o~ between sun and shade leaves and the simulated sunfleck where then compared as a function of C~i~:C~a~.

##Biochemical parameters of photosynthesis
Photosynthetic CO~2~ response (AC~i~) curves were measured at 25 &deg;C for one sun and shade leaf for each chamber prior to the initiation of the drought treatment. Each AC~i~ curve began at the reference [CO~2~] of 400 $\mu$l l^-1^ and then consisted of 12 additional steps from [CO~2~] of 50 to 1800 $\mu$l l^-1^ at 25 &deg;C at saturating light (1800 $\mu$mols m^-1^ s^-1^). From these curves the photosynthetic parameters, J~max~ and Vc~max~, were quantified using the biochemical model of [@farquhar1980biochemical] and fit with the 'plantecophys' package [@Duursma2014] in R [@RDevelopmentCoreTeam2011].

##Leaf chemistry and hydraulic parameters
Following gas exchange measurements each leaf was collected, measured for leaf water potential (explained below), scanned for leaf area, dried to constant mass and weighed. These leaves were then milled and analyzed for leaf N content and $\delta$^13^C. Leaf samples were analysed on a Delta V Advantage coupled to a Flash HT and Conflo IV (Thermo Fisher Scientific, Bremen, Germany) in dual-reactor setup. Samples were flash combusted at 1000&deg;C and converted to CO~2~ and N~2~ and then subjected to stable isotope ratio mass spectrometry. Leaf N is reported on an area basis (N~a~) and isotopic signatures are reported relative to the VPDP scale.

\
Predawn ($\Psi_{pd}$) and midday ($\Psi_l$) leaf water potentials (MPa) were measured for sun and shade leaves during each gas exchange campaign using a PMS 1505D pressure chamber (PMS Instruments, Albany, OR, USA). The leaf closest to the leaf used for gas exchange was sampled for measurement of $\Psi_{pd}$. Predawn leaf water potential was measured before sunrise on the same day as gas exchange measurements. Leaves used for gas exchange were immediately sampled for $\Psi_l$ once measurements were completed. All leaves were detached and immediately stored inside foil covered bags before water potential measurements were performed. Leaf water potential and transpiration (E, mmol^-2^ s^-1^) from gas exchange were then used to calculate leaf-specific hydraulic conductance (K) through the equation:

@@

## Photosynthetic capacity and leaf photosynthesis rates 
The photosynthetic parameters J~max~ and Vc~max~ were higher in sun compared to shade leaves (Table 1), as estimated from AC~i~ curves from each chamber (Figure `r acifig`a). Within leaf types, no effect of the warming treatment was detected on either parameter. Within individual chambers Vc~max~ was positively related to leaf N~a~ across leaf types and temperature treatments (P = 0.01, Figure `r acifig`b). As g~m~ (results below) is implicit in the estimates of J~max~ and Vc~max~ with conventional AC~i~ curves, we also simulated AC~c~ curves to determine if treatment differences in AC~i~ parameters where instead the result of differences in g~m~. When accounting for the effect of g~m~, chloroplastic Vc~max~ and J~max~ were still higher in sun leaves (Figure `r accfig`).

\
Mean A was significantly higher in sun compared to shade leaves (+23%), when measured at their local light environment (Table 2). Additionally, leaf N~a~ was positively related to A across gas exchange campaigns and leaf types (P < 0.001, Figure `r acifig`c). Following an increase light intensity, A of shade leaves was significantly greater than both shade leaves at low light and sun leaves (P < 0.001). No effect of the warming treatment was detected on A of sun leaves or shade leaves at low or high light. Photosynthesis within leaf types and treatments was similar through time and across the range of leaf temperatures measured (Figure `r get1fig`a).

## Stomatal conductance and leaf water use efficiency
On average, g~s~ was 18% higher in shade compared to sun leaves under their local light environment (Table 2). In sun leaves, A was positively correlated with g~s~, however, g~s~ and A were not correlated in shade leaves under low light (Figure `r agmgsfig`a). Following increased PPFD, g~s~ of shade leaves was significantly greater than both shade leaves at low light and sun leaves (Figure `r gebarfig`a). With increased PPFD, increases in A and g~s~ in shade leaves were positively correlated  as observed in sun leaves.

\
Leaf ITE was significantly greater in sun leaves than in shade leaves at low light (+21%, P = 0.001). Following an increase in  PPFD, ITE of shade leaves did not differ from shade leaves at low light and was still significantly lower than sun leaves (P < 0.001). Instantaneous transpiration efficiency in sun leaves was reduced in the warming treatment but no effect was noted in shade leaves with low or high light (Table 2). For all leaf types and light treatments there was a strong response of ITE to VPD and individual data points broadly corresponded to response curves from the optimal WUE model (Figure `r waterusefig`a). The mean estimated g~1~ for sun leaves was 1.51&pm;0.11 and for shade leaves with low and high light was 2.59&pm;0.12 and 2.74&pm;0.04. Within leaf types and light treatments the response of VPD to leaf temperature was similar across all measurement campaigns (Figure `r get2fig`a)

\
Bulk-leaf $\delta$^13^C, as an index of integrated water use efficiency [@marshall2007sources], was significantly lower in shade leaves compared to sun leaves by ca. 2&permil; (Table 1). No effects of the warming treatment on leaf $\delta$^13^C were detected. Leaf $\delta$^13^C and N~area~ were positively correlated for all leaves (P<0.001, Figure `r waterusefig`b), with less negative $\delta$^13^C and higher N investment in sun leaves.

@@

\
It is possible that reducing stomatal response time, by sustaining higher g~s~, is a strategy to take advantage of high light quickly in shade leaves [@tausz2005dynamic]. Evidence from this study supports this hypothesis, as shade leaves increased A equivalent or even outperforming sun leaves under identical light intensity. Transpiration-induced cooling in shade leaves, by keeping stomata open, has also been suggested as an effective strategy to reduce sunfleck induced thermal load [@schymanski2013stomatal]. This is because rapid increases in leaf temperature with sunflecks have been shown to inhibit C gain [@leakey2003high]. However, this response likely occurs at already high air temperatures and may not explain the observed g~s~ in shade leaves across the large range of temperatures included in this study. How prevalent each of these strategies are within tree canopies is still unknown, as empirical studies assessing photosynthetic responses to sunflecks generally focus on seedlings [@kuppers1993leaf;@pepin1997rates;@leakey2002relative] and understory plants [@chazdon1991importance; @allen2000stomatal; @brantley2009contribution]. Thus, our findings highlight a critical need for empirical measurements of shade leaves under dynamic light environments in order to accurately scale C gain from leaf to canopy [see @de1997simple].

\
We found that A and g~m~ scaled positively across leaf types and, surprisingly, increased proportionately when light intensity was increased in shade leaves.  Research has suggested that aquaporins can facilitate increases in the CO~2~ permeability of the cell membranes resulting in rapid modulation of g~m~ [@hanba2004overexpression; @heinen2009role; @li2014plant]. This provides a potential explanation for the observed rapid increases in g~m~ but the impacts of aquaporins on g~m~ are untested in leaves of tree species. Our findings support a growing wealth of evidence that g~m~ is highly variable and can respond to environmental variables [@flexas2008mesophyll]. Here we provide empirical data showing g~m~ not only varies within a canopy but the up-regulation of g~m~ plays a critical role in the photosynthetic response of shade leaves to sunflecks.

@@

\
**Figure 4**. The response of A to g~s~ (a) and g~m~ (b) for sun leaves and shade leaves at both low and high light with elevated and ambient temperature treatments. Lines represent either smoothed regressions from a generalized additive model fit (a) or linear model fits (b). Grey areas are 95% confidence intervals from the mean.

\
**Figure 5**. The mean &pm; 1 standard error of g~s~ (a), g~m~ (b) and A (c) of sun leaves and shade leaves at both low and high light.

\
**Figure 6**. (a) Response of ITE to VPD for sun leaves and shade leaves at both low and high light with elevated and ambient temperature treatments. (b) The relationship between leaf $\delta$^13^C and leaf N~a~ for sun leaves and shade leaves at low light. For (a) VPD is the leaf to air pressure difference inside the gas exchange cuvette and lines represent predictions from the optimal ITE model with a g~1~ value for each leaf type and treatment. For (b) the dashed line represents the significant linear model fit for all leaves with a marginal and conditional R^2^ of 0.41 and 0.45, respectively.

\
**Figure 7**. Relationship between the observed discrimination of ^13^CO~2~ and measured C~i~/C~a~ for sun leaves and shade leaves at both low and high light. The solid line represents the theoretical line for C3 plants from Evans et al. [-@evans1986carbon].

>>>>

\
During A, CO~2~ diffuses from the atmosphere through stomata (g~s~), intercellular air space and into the chloroplast for fixation by Rubisco (mesophyll conductance, g~m~) [@tazoe2011using]. Based on optimality theory, stomatal conductance should be efficiently distributed within a canopy to utilize supplies of light, N and water to maximize A [@peltoniemi2012co]. This is because stomata are hypothesized to exhibit an optimal behaviour to maximize C gain while simultaneously minimizing water loss through transpiration [@cowan1977stomatal]. Mesophyll conductance can also impose limitations on A as large g~s~ [@warren2008stand; @ubierna2011estimation], reducing the efficiency of leaf N use in A [@niinemets2007photosynthesis] if g~m~ constrains CO~2~ supply to the chloroplast. Part of the variation in photosynthetic capacity between sun and shade leaves has been proposed to be due to differences in g~m~ [@piel2002effect; @warren2007internal], yet the trade-offs that constrain this diffusion pathway are yet to be explicitly quantified. Stomatal and mesophyll conductance should not be considered independent of each other [@griffiths2013mesophyll], but a lack empirical data currently hinders are ability to interpreting their coupled responses to A across sun and shade leaves.

\
Additionally, assessing shade leaf behaviour is made difficult with accounting of short term light fluctuations within a canopy, via sunflecks. How shade leaves utilize sunflecks for short term C gain depends on the combined response time of g~s~ and g~m~ and the underlying photosynthetic biochemistry acclimated to a low light environment [@pearcy1990sunflecks;@tausz2005dynamic]. For example, at short timescales the utilization of sunflecks is limited by delayed responses of stomata, effectively limiting the maximum assimilation rate that can be achieved [@vico2011effects; @way2012sunflecks]. Mesophyll conductance has been shown to respond to environmental factors (e.g. CO~2~, temperature or vapor pressure deficit) across short time scales, possibly faster than g~s~ [@flexas2008mesophyll and references therein], yet short term response to light availability are unclear. For example, g~m~ was found to be independent of light intensity in wheat leaves [@tazoe2009light] but was responsive to light in tobacco [@flexas2007rapid]. Anatomical parameters which regulate g~m~ with changing irradiance such as chloroplast surface area [@terashima2006irradiance] and mesophyll thickness [@boardman1977comparative] are also unlikely to adjust during short light fluctuations. The physiological behaviour of shade leaves to maximize C gain must be assessed as both a degree of acclimation to local irradiance and as a potential response to transitory light availability.

@@

\
In this study we use *Eucalyptus tereticornis* Sm. trees, planted in climate controlled whole tree chambers with ambient and elevated temperature (ambient +3&deg;C) treatments, to empirically evaluate the distribution of resources and leaf physiological behaviour of sun and shade leaves. Our hypotheses are as follows:

@@

\
Before seedlings were planted into each chamber they were maintained under well-watered conditions in 25 L pots and kept inside each chamber. This allowed for seedlings to gain sufficient size before planting while also allowing them to acclimate to chamber temperature treatments. Seedlings were planted into each chamber after mean seedling height reached 100 cm. The top soils at this site, used in both pots and chambers, are an alluvial formation of low-fertility sandy loam soils with low organic matter and low water holding capacity. A root exclusion barrier extended from chamber walls to the hard layer (ca. 1 m) and roots were allowed to grow freely below the barrier. Leaf gas exchange measurements were initiated in October 2013 when trees had both ample height growth and canopy development for realistic canopy light gradients to be measured. At this point, trees under AT treatment had a mean diameter of 28.2&pm;1.1 mm, height of 348&pm;15.1 cm and an leaf area of 3.9&pm;0.1 m^2^. For ET treatments, trees had a mean diameter of 34.1&pm;2.1 mm, height of 418.3&pm;23.1 cm and an leaf area of 6.2&pm;0.2 m^2^.

##Leaf gas exchange, online carbon isotope discrimination and mesophyll conductance
Leaf gas exchange measurements were performed six times, beginning in October 2013 and monthly from December 2013 to April 2014. Measurements were taken on a representative sun and shade leaf for each tree during each measurement campaign. The newest fully expanded leaf from the branch apex was chosen for gas exchange measurements and sun leaves were measured in the upper third of the canopy. Here, shade leaves are defined as inner-canopy leaves developing in a low light environment. In order to minimize any confounding effects of leaf age, shade leaves were measured first in the lower canopy then gradually higher during each campaign. As shade leaves most likely developed slower this assured that older leaves in the lower canopy were avoided.

\
Prior to gas exchange measurements photosynthetic photon flux density (PPFD) was recorded as a point measurement at the individual leaf level and a spatially averaged measurement at the canopy position for each selected leaf.  A hand-held quantum sensor (LI-COR, Lincoln, NE, USA) was used to record leaf level PPFD to ensure that chosen leaves were positioned in the desired light environment, either full sun or full shade. A ceptometer (AccuPAR LP-80, Decagon Devices, Pullman, USA) was then used to measure integrated PPFD at the canopy height of each chosen leaf type. Each ceptometer reading integrated an array of 80 sensors over a total length of 84 cm. Five ceptometer readings were recorded within the canopy at the same height and close to each selected leaf. The mean of these readings was assumed to represent the local light environment of sun and shade leaves for each tree. All measurements of PPFD and gas exchange were performed on sunny days between 10:00-14:30 h.

\
Leaf level gas exchange was measured with a standard (2 x 3 cm) leaf cuvette using a portable gas exchange system (LI-6400XT, LI-COR, Lincoln, NE, USA).  This system was coupled with a tunable diode laser (TDL; TGA100,Campbell Scientific, Inc., Logan, UT, USA) for concurrent measurements of online C isotope discrimination. The CO~2~ in the leaf cuvette was set at ambient atmospheric [CO~2~] (400 ppm) with a flow rate of 200 $\mu$mol s^-1^. Two identical gas exchanges systems were run simultaneously, one in each of a randomly chosen WTC for each temperature treatment. Leaf temperatures were controlled at the current AT or ET WTC air temperature. PPFD in the cuvette was set to match the individual light environment of each leaf type (explained above). Sun flecks were simulated for shade leaves by increasing the leaf cuvette PPFD (LI-COR red/blue light source) to match the light environment of the full sun leaf in the same tree. The maximum sunfleck response was recorded once CO~2~ and water vapor fluxes re-stabilized in the leaf cuvette (ca. 25 min).

\
Once CO~2~ and water vapor flux values were stable for each leaf measurement, the sample and reference gas lines were diverted to the TDL via T-junctions inserted into the reference gas tube and match valve outlet of the LI-6400. The gas streams were dried by passing through napion gas dryers in the respective gas lines, and then ^12^CO~2~ and ^13^CO~2~ concentrations were measured for each gas stream by the TDL. Reference, sample and two calibration gases were run on alternating 80 s loops (20 s each), one for each AT and ET leaf at a matched canopy position, for a total of 12 min. This allowed for 4-5 measurements per leaf and data were averaged over the last 10 s of reference line and samples line gas streams for calculations. The two calibration gases were drawn from compressed air tanks (330 and 740 ppm CO~2~) in order to correct for gain drift of the TDL on each measurement cycle. Photosynthesis, g~s~, transpiration, VPD and intercellular [CO~2~] (C~i~) were auto-logged every 15 s for each gas exchange system over the same 12 min interval.

\
Using online C isotope discrimination measurements, the difference  between the observed discrimination and what is predicted for light saturated gas exchange is proportional to g~m~ [@griffiths2013mesophyll]. First, leaf C isotope discrimination was calculated by comparing the isotopic composition of the reference gas entering the leaf cuvette ($\delta$^13^C~e~) with the sample gas ($\delta$^13^C~o~) such that:

@@

and $\xi$ is the ratio of the CO~2~ entering the well mixed leaf cuvette to the CO~2~ draw down in the gas stream by the leaf.

@@

\
(7)
$$
\begin{aligned}
\Delta_e = &\frac{1+t}{1-t}\left(\frac{eR_d}{(A+R_d)C_a}(C_i-\Gamma_*)\right)
\end{aligned}
$$

@@

where C~s~ is the CO~2~ partial pressure at the leaf surface, a~b~ is the fractionation from boundary layer diffusion (2.9&permil;) and a is the fractionation due to diffusion in air (4.4&permil;) [@evans1986carbon]. C~a~ and C~i~ are the atmospheric and intercellular partial pressures and $\Gamma_{*}$ is the compensation point in the absence of mitochondrial respiration in the light (R~d~). In this experiment both $\Gamma_{*}$ and R~d~ were derived using a standard Arrhenius function with parameters for *Eucalyptus globulus* from [@crous2012light]. The ternary effect corrections (t) are described by:

@@

where C~c~ is the chloroplast CO~2~ partial pressure. Once g~m~ was calculated C~c~ and the drawdown of CO~2~ from the intercellular air spaces to the site of carboxylation were then estimated using Equation 11. Examples of this approach to measure gas exchange and C isotope discrimination are presented in Evans and Von Caemmerer [-@evans2013temperature]. The variation in $\Delta$~o~ between sun and shade leaves and the simulated sunfleck where then compared as a function of C~i~:C~a~.

##Biochemical parameters of photosynthesis
Photosynthetic CO~2~ response (AC~i~) curves were measured at 25 &deg;C for one sun and shade leaf for each WTC prior to the initiation of the drought treatment. Each AC~i~ curve began at the reference [CO~2~] of 400 $\mu$l l^-1^ and then consisted of 12 additional steps from [CO~2~] of 50 to 1800 $\mu$l l^-1^ at 25 &deg;C at saturating light (1800 $\mu$mols m^-1^ s^-1^). From these curves the photosynthetic parameters, J~max~ and Vc~max~, were quantified using the biochemical model of [@farquhar1980biochemical] and fit with the 'plantecophys' package [@Duursma2014] in R [@RDevelopmentCoreTeam2011].

##Leaf chemistry and hydraulic parameters
Following gas exchange measurements each leaf was collected, measured for leaf water potential (explained below), scanned for leaf area, dried and weighed. These leaves were then milled and analyzed for leaf N content and $\delta$^13^C. Leaf samples were analysed on a Delta V Advantage coupled to a Flash HT and Conflo IV (Thermo Fisher Scientific, Bremen, Germany) in dual-reactor setup. Samples were flash combusted at 1000&deg;C and converted to CO~2~ and N~2~ and then subjected to stable isotope ratio mass spectrometry. Leaf N is reported on an area basis (N~a~) and isotopic signatures are reported relative to the VPDP scale.

\
Predawn ($\Psi_{pd}$) and midday ($\Psi_l$) leaf water potentials (MPa) were measured for sun and shade leaves during each gas exchange campaign using a PMS 1505D pressure chamber (PMS Instruments, Albany, OR, USA). The leaf closest to the leaf used for gas exchange was sampled for measurement of $\Psi_{pd}$. Predawn leaf water potential was measured before sunrise on the same day as gas exchange measurements. Leaves used for gas exchange were immediately sampled for $\Psi_l$ once measurements were completed. All leaves were detached and immediately stored inside foil covered bags before water potential measurements were performed. Leaf water potential and transpiration (E, mmol^-2^ m^-2^ s^-1^) from gas exchange were then used to calculate leaf-specific hydraulic conductance (K) through the equation:

@@

## Photosynthetic capacity and leaf photosynthesis rates 
The photosynthetic parameters J~max~ and Vc~max~ were higher in sun compared to shade leaves (Table 1), as estimated from AC~i~ curves (Figure `r acifig`a). Within leaf types, no effect of the warming treatment was detected on either parameter. Among the sampled leaves, Vc~max~ was positively related to leaf N~a~ across leaf types and temperature treatments (P = 0.01, Figure `r acifig`b). As g~m~ (results below) is implicit in the estimates of J~max~ and Vc~max~ with conventional AC~i~ curves, we also simulated AC~c~ curves to determine if treatment differences in AC~i~ parameters where instead the result of differences in g~m~. When accounting for the effect of g~m~, chloroplastic Vc~max~ and J~max~ were still higher in sun leaves (Figure `r accfig`).

\
Mean A was significantly higher in sun compared to shade leaves (+23%), when measured at their local light environment (Table 2). Additionally, leaf N~a~ was positively related to A across gas exchange campaigns and leaf types (P < 0.001, Figure `r acifig`c). Following an increase light intensity to match high-light conditions, A of shade leaves increased to values significantly greater than sun leaves at high light (P < 0.001). No effect of the warming treatment was detected on A of sun leaves measured at high light or shade leaves at either low or high light. Photosynthesis within leaf types and treatments was similar through time and across the range of leaf temperatures measured (Figure `r get1fig`a).

## Stomatal conductance and leaf water use efficiency
On average, g~s~ was 18% higher in shade compared to sun leaves under their local light environment (Table 2). In sun leaves, A was positively correlated with g~s~, however, g~s~ and A were not correlated in shade leaves under low light (Figure `r agmgsfig`a). Following increased PPFD, g~s~ of shade leaves was significantly greater than both shade leaves at low light and sun leaves, pooled across all measurement dates (Figure `r gebarfig`a). With increased PPFD, increases in A and g~s~ in shade leaves were positively correlated  as observed in sun leaves.

\
Measured under ambient light and temperature conditions leaf ITE was significantly greater in sun leaves than in shade leaves at low light (+21%, P = 0.001). Following an increase in  PPFD to high-light conditions, ITE of shade leaves did not differ from shade leaves at low light and was still significantly lower than sun leaves (P < 0.001). Instantaneous transpiration efficiency in sun leaves was reduced in the warming treatment but no warmding effect was detected in shade leaves measured at low or high light (Table 2). For all leaf types and light treatments there was a strong response of ITE to VPD and individual data points broadly corresponded to response curves from the optimal WUE model (Figure `r waterusefig`a). The mean estimated g~1~ for sun leaves was 1.51&pm;0.11 and for shade leaves with low and high light was 2.59&pm;0.12 and 2.74&pm;0.04. Within leaf types and light treatments the response of VPD to leaf temperature was similar across all measurement campaigns (Figure `r get2fig`a).

\
Bulk-leaf $\delta$^13^C, as an index of integrated water use efficiency [@marshall2007sources], was significantly lower in shade leaves compared to sun leaves by ca. 2&permil; (Table 1). No effects of the warming treatment on leaf $\delta$^13^C were detected. Leaf $\delta$^13^C and N~area~ were positively correlated for all leaves (P<0.001, Figure `r waterusefig`b), with less negative $\delta$^13^C (higher water-use efficiency) and higher N investment in sun leaves.

@@

\
It is possible that reducing stomatal response time, by sustaining higher g~s~, is a strategy to take advantage of high light quickly in shade leaves [@tausz2005dynamic]. Evidence from this study supports this hypothesis, as shade leaves increased A equivalent or even outperforming sun leaves under identical light intensity. Transpiration-induced cooling in shade leaves, by keeping stomata open, has also been suggested as an effective strategy to reduce sunfleck induced thermal load [@schymanski2013stomatal]. This is because rapid increases in leaf temperature with sunflecks have been shown to inhibit C gain [@leakey2003high]. However, this response likely occurs at already high air temperatures and may not explain the observed g~s~ in shade leaves across the large range of temperatures included in this study. How prevalent each of these strategies are within tree canopies is still unknown, as empirical studies assessing photosynthetic responses to sunflecks generally focus on seedlings [@kuppers1993leaf;@pepin1997rates;@leakey2002relative] and understory plants, often in deep shade [@chazdon1991importance; @allen2000stomatal; @brantley2009contribution]. Thus, our findings highlight a critical need for empirical measurements of shade leaves under dynamic light environments in order to accurately scale C gain from leaf to canopy [see @de1997simple].

\
We found that A and g~m~ scaled positively across leaf types and, surprisingly, increased rapidly (within minutes) and proportionately when light intensity was increased in shade leaves.  Research has suggested that aquaporins can facilitate increases in the CO~2~ permeability of the cell membranes resulting in rapid modulation of g~m~ [@hanba2004overexpression; @heinen2009role; @li2014plant]. This provides a potential explanation for the observed rapid increases in g~m~ but the impacts of aquaporins on g~m~ are untested in leaves of tree species. Our findings support a growing wealth of evidence that g~m~ is highly variable and can respond to environmental variables [@flexas2008mesophyll]. Here we provide empirical data showing g~m~ not only varies within a canopy but the up-regulation of g~m~ plays a critical role in the photosynthetic response of shade leaves to sunflecks.

@@

\
**Figure 4**. The response of A to g~s~ (a) and g~m~ (b) for sun leaves measured at high light and shade leaves measured at both low and high light under their respective elevated and ambient temperature treatments. Lines represent either smoothed regressions from a generalized additive model fit (a) or linear model fits (b). Grey areas are 95% confidence intervals from the mean.

\
**Figure 5**. The mean &pm; 1 standard error of g~s~ (a), g~m~ (b) and A (c) of sun leaves and shade leaves at both low and high light pooled across six measurement dates.

\
**Figure 6**. (a) Response of instantaneous transpiration efficiency (ITE) to VPD for sun leaves and shade leaves at both low and high light with elevated and ambient temperature treatments. (b) The relationship between leaf $\delta$^13^C and leaf N~a~ for sun leaves at high light and shade leaves at low light. For (a) VPD is the leaf to air pressure difference inside the gas exchange cuvette and lines represent predictions from the optimal ITE model with a g~1~ value for each leaf type and treatment. For (b) the dashed line represents the significant linear model fit for all leaves with a marginal and conditional R^2^ of 0.41 and 0.45, respectively.

\
**Figure 7**. Relationship between the observed discrimination of ^13^CO~2~ measured during photosynthesis and measured C~i~/C~a~ for sun leaves measured at hight light and shade leaves measured at both low and high light. The solid line represents the theoretical line for C3 plants from Evans et al. [-@evans1986carbon].
